---
title: "Introduction to penguinglyphs"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
bibliography: ["references.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Introduction to penguinglyphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

In data visualization, a "glyph" is a small graphical object, such as a shape, icon, or marker, that represents a single observation. Glyphs are used to visualize multidimensional data by encoding different variables into various visual attributes like shape, size, color, and orientation. The use of "glyphs" to represent multivariate data has a long history [@Ward2008]. One of the early uses, by Edgar Anderson [-@Anderson:1957], used circular glyphs with rays to study botanical species, including the famous Iris Flowers [@Kleinman:2002].
More recently, @Chernoff:1973 introduced the use of schematic faces that could represent up to 15 data variables
simultaneously.

The `penguinglyphs` package is an experiment in the use of glyphs to represent the observations in
the datasets on Penguins in Antartica, now in the R datasets `datasets::penguins`.
The first step was to see how Penguin features, like body mass, flipper length and the length and depth of
their bills could be represented **parametrically**, and thus encoded in ways to permit drawing a penguin observation using R graphics like `polygon()`, `segments()` and so forth.


## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("friendly/penguinglyphs")
```

```{r setup}
library(penguinglyphs)
```

## Visual Encoding

The package uses the following mappings:

- **Bill length** → horizontal extent of the bill
- **Bill depth** → vertical thickness of the bill
- **Flipper length** → length of the flippers
- **Body mass** → area of the body
- **Species** → body color (Adelie=orange, Chinstrap=purple, Gentoo=green)
- **Sex** → eye shape (angular for males, round for females)

Each measurement is converted to a scale factor between 0.7 and 1.3, according to where it falls in the range
of that variable in the full `penguins` dataset. The vignette *How to Draw a Penguin* shows how the glyph is
constructed, and what each of these features looks like over its range.

## Basic Usage

First, let's load the Palmer Penguins dataset:

```{r}
# Load penguin data
data(penguins, package = "datasets")

# Look at the structure
head(penguins)
```

### Creating a Grid of Glyphs

The main function is `penguin_glyphs()`, which creates a grid display, 5 columns wide by default.
Here, I show the first five penguins of each species.

```{r}
#| label: fig-peng15
#| fig-width: 8
#| fig-height: 6
#| fig-cap: Glyphs for the first five penguins in each species
# Visualize first 5 penguins in each species
which <- outer(1:5, c(0, 152, 277), FUN ="+") |> c()
penguin_glyphs(penguins[which,])
```

What happened with case 4? All of its measurements are missing (`NA`), so the glyph is drawn only as a dashed outline.
Henceforth, I'll work with the subset, `peng`, of the 333 penguins that have no missing values.

```{r}
peng <- na.omit(penguins)
rownames(peng) <- NULL
nrow(peng)
```

The row names are used as the labels in the glyphs. `na.omit()` keeps the original ones, which then have gaps where
penguins were dropped; resetting them numbers the complete cases from 1 to 333.
This matters for identifying particular penguins. `peng` is then the same as the dataset `heplots::peng`, apart from the
names of some variables, so the case numbers shown here are the same as those in analyses of that dataset, as in
@Friendly:2024:VisMLM.


### Random Sample

```{r fig.width=8, fig.height=6}
set.seed(42)
sampled_rows <- sample(1:nrow(peng), size = 20)
penguin_glyphs(peng[sampled_rows, ], main = "Random Sample of Penguins")
```

## Drawing Individual Penguins

For more control, you can draw individual penguins with `draw_penguin()`:

```{r fig.width=6, fig.height=4}
# Create plot area
plot(1, xlim=c(0,4), ylim=c(0,2), type="n", asp=1,
     xlab="", ylab="", main="Species Comparison")

# Draw one of each species; size is the height of the glyph, in inches
draw_penguin(1, 1, species="Adelie", sex="male", id="Adelie", size = 1.5)
draw_penguin(2, 1, species="Chinstrap", sex="female", id="Chinstrap", size = 1.5)
draw_penguin(3, 1, species="Gentoo", sex="male", id="Gentoo", size = 1.5)
```

## Glyphs in Other Plots

A glyph is sized in inches rather than in the units of the axes, so it has the same shape and size in any plot,
in the same way as a plotting symbol.
`penguin_points()` uses this to add a glyph for each row of a data frame at given coordinates in an existing plot,
as `points()` does for ordinary symbols. `penguin_legend()` adds a legend for the species.

Here, the glyphs for a sample of 40 penguins are the points in a scatterplot of body mass against flipper length.
The glyphs repeat what the axes show, in that penguins toward the right have longer flippers and those toward the top have
larger bodies, but they also show the bill and the sex of each penguin.

```{r}
#| label: fig-scatter
#| fig-width: 8
#| fig-height: 6
#| fig-cap: Penguin glyphs as the point symbols in a scatterplot
set.seed(42)
show <- sample(nrow(peng), 40)
samp <- peng[show, ]

plot(body_mass ~ flipper_len, data = samp, type = "n",
     xlab = "Flipper length (mm)", ylab = "Body mass (g)")
penguin_points(samp$flipper_len, samp$body_mass, samp)
penguin_legend("topleft", samp$species)
```

The coordinates need not be variables in the data. A more useful display puts the glyphs in the space
of the first two principal components, so that position summarizes all four measurements, and the glyph shows what
lies behind it.

```{r}
#| label: fig-pca
#| fig-width: 8
#| fig-height: 6
#| fig-cap: Penguin glyphs in the space of the first two principal components
pca <- prcomp(peng[, 3:6], scale. = TRUE)

plot(pca$x[, 1:2], type = "n", asp = 1)
penguin_points(pca$x[show, 1], pca$x[show, 2], samp)
penguin_legend("top", samp$species, horiz = TRUE)
```

## Finding Patterns

### Examining Outliers

Unusual multivariate cases are easily identified using Mahalanobis $D^2$. Here, I find the three most noteworthy observations
in the total sample (ignoring `species`).

```{r}
library(heplots)
DSQ <- heplots::Mahalanobis(peng[, 3:6])
outliers <- order(DSQ, decreasing = TRUE)[1:3] |>
  print()
```

You can see them in a $\chi^2$ QQ plot of the quantiles of the $D^2$ values against what they would be in a multivariate normal
distribution for four variables.

```{r}
clr <- c(Adelie = "#F37A00", Chinstrap = "#6A3D9A", Gentoo = "#33a02c")
pch <- c(19, 17, 15)   # ggplot symbol defaults for a factor

out <- heplots::cqplot(peng[, 3:6], 
   id.n = 3,
   col = clr[peng$species],
   pch = pch[peng$species],
   ref.col = "grey",
   what = "Penguin numeric variables",
   cex.lab = 1.25)
out
```


What makes these penguins unusual? Their glyphs are labeled with the same case numbers as in the QQ plot.

```{r fig.width=6, fig.height=3}
# Look at some notable peng
penguin_glyphs(peng[outliers,], ncol=3, main = "Notable Penguins")
```

It also helps to see where they are in relation to the rest. Here, all the penguins are shown as small points
in plots of the principal components, and only the three outliers are drawn as glyphs.

The plot of the first two components, which account for most of the variance, is the one usually looked at.
But none of the three is at all unusual there. Multivariate outliers tend to show up instead in the dimensions
with the *smallest* variance [@Friendly:2024:VisMLM]: these are the directions in which the data vary least,
so a penguin that departs from the pattern of correlations among the variables stands out.
In the plot of the last two components, all three are on the periphery: penguins 283 and 10 are at opposite ends of the third dimension,
and 35 is at the top of the fourth.

```{r}
#| label: fig-pca-outliers
#| fig-width: 10
#| fig-height: 5.5
#| fig-cap: "The three most unusual penguins, shown as glyphs in plots of the first two principal components (left) and the last two (right)."
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (dims in list(1:2, 3:4)) {
  plot(pca$x[, dims], asp = 1, pch = 16, cex = 0.7,
       xlim = extendrange(pca$x[, dims[1]], f = 0.15),   # room for the glyphs
       col = penguin_colors()[as.character(peng$species)],
       main = paste("Dimensions", dims[1], "and", dims[2]))
  penguin_points(pca$x[outliers, dims[1]], pca$x[outliers, dims[2]], peng[outliers, ],
                 size = 0.7, id = TRUE)
}
penguin_legend("bottomright", peng$species)
par(op)
```

### Visual Lineup

A "line-up" test [@Buja-etal2009]
has been proposed as a **human significance test**: Can an observer spot a difference that rejects a null hypothesis?
This is more a test of a data visualization per se than of the human observer. 



Here, the idea is turned around, to ask whether the unusual penguins can be picked out from their glyphs alone.
The display shows 17 penguins chosen at random, together with the three outliers found above, in a random order.

```{r fig.width=8, fig.height=6}
# Create a lineup for visual inference
set.seed(123)
cast <- c(sample(1:nrow(peng), size = 17), outliers)
lineup <- sample(cast, size = length(cast))
penguin_glyphs(peng[lineup,], main = "Can you spot the outliers?")
```

The outliers are peng `r paste(rownames(peng)[outliers], collapse = ", ")`. If you can't find them
without the labels, that says something about how well these glyphs convey what makes a penguin unusual in four dimensions,
which need not be an extreme value on any one of them, as the principal component plots showed.

## Customization

### Legend Placement

The legend is placed outside the grid of glyphs, at the top left by default. You can put it on another side,
or lay it out horizontally.

```{r fig.width=8, fig.height=6}
penguin_glyphs(peng[sampled_rows, ], 
               main = "Custom Legend",
               legend = list(loc = "bottom", horiz = TRUE))
```

### Scaling

By default, the size of each feature reflects where a penguin falls in the range of the *whole* dataset,
so a penguin looks the same whichever others it is shown with. The `ref` argument changes the reference data.
With `ref = data`, the features are scaled to the ranges in just the penguins being plotted,
which exaggerates the differences among them. Compare the three outliers, scaled in each way.

```{r fig.width=6, fig.height=3}
penguin_glyphs(peng[outliers,], ncol = 3, main = "Scaled to all peng")
```

```{r fig.width=6, fig.height=3}
penguin_glyphs(peng[outliers,], ncol = 3, main = "Scaled to these three",
               ref = peng[outliers,])
```

### Label Size

`cex.lab` controls the size of the labels in the glyphs.

```{r fig.width=8, fig.height=6}
penguin_glyphs(peng[sampled_rows, ], cex.lab = 0.7, main = "Smaller labels")
```

### Different Grid Layouts

```{r fig.width=10, fig.height=4}
# Fewer columns for a wider display
penguin_glyphs(head(peng, 15), 
               ncol = 8, 
               main = "Wide Layout")
```

## Interpreting the Glyphs

When looking at penguin glyphs, the features are not all equally easy to read:

1. **Species** is identified immediately, by color.
2. **Body mass** is the most conspicuous of the measurements: a heavier penguin has a larger body.
3. **Bill proportions**: length and depth are read together, as long, thin bills vs. short, deep bills.
4. **Flipper length**: longer flippers extend further from the body, but the differences are modest.
5. **Sex**: the difference in eye shape is consistent, but subtle, and hard to see when the glyphs are small.

In the displays above, this is enough to see species-specific morphology, such as the large bodies, long flippers
and thin bills of the Gentoos. Differences among the penguins of one species, including those between males and females,
are harder to see.

## Conclusion

The `penguinglyphs` package is an experiment in using a glyph that is a schematic picture of the thing that was measured,
so that the mapping of variables to visual features does not have to be learned.
It provides a grid display of glyphs for a set of penguins, and glyphs that can be used as the points in any other plot.

Some possible uses are:

- Exploratory data analysis, particularly in combination with a dimension-reduction plot
- Showing what makes multivariate outliers unusual
- Teaching ideas of visual encoding, and how to evaluate one

Whether such glyphs work better than more abstract ones, like stars or Chernoff faces, remains to be tested.

## References
